Quality Criteria for SKOS Vocabularies

### 2012-03-22 criteria.tex
finished criteria

The following issues primarily focus on possible problems that might arise when human users work with a SKOS vocabulary or interact with a system that uses the vocabulary for, e.g., search and retrieval purposes.

@@

The commonality of the following issues is that they can affect applications areas that exploit the underlying graph structure of a SKOS vocabulary.

@@

%Since all function operate on the entailed RDF graph, this algorithm implicitly also considers \texttt{skos:narrower} relations. CM: ist eigentlich aus def1 klar denke ich

@@

% Origin and rationale
The ISO/DIS 25964-1 standard \cite{ISO25964-1:2011} suggests that terms that share a common broader term but do not have an overlapping meaning, should not be related associatively. This is advocated by \cite{Hedden2010,Aitchison2000} the latter also mentioning ``the risk that thesaurus compilers may overload the thesaurus with valueless relationships'', having a negative effect on precision.
% Function defintion
The function \textit{var} identifies a set of pairs of conceptual resources that share the same broader or narrower concept while also being associatively related by the property \texttt{skos:related}.

\paragraph{QI 9: Solely Transitively Related Concepts}
% Origin and rationale
Two concepts that are \textbf{explicitly} related by \texttt{skos:broaderTransitive} and/or \texttt{skos:narrowerTransitive} can be regarded a quality issue because, according to \cite{SkosReference2008}, they are ``not used to make assertions''. Transitive hierarchical relations in SKOS are meant to be infered by the vocabulary \textbf{consumer}, which is reflected in the SKOS schema by, e.g., \texttt{skos:broader} being a subproperty of \texttt{skos:broaderTransitive}. 
% Function definition
This fuction \textit{strc} iterates over all pairs of concepts $C \times C$ that are directly related by the mentioned transitive properties, not by one of their subproperties.

\paragraph{QI 10: Omitted Top Concepts}
% Origin and rationale
The SKOS language provides concept schemes, which are a facility for grouping related concepts. In order to provide entry points to such a group of concepts, one or more concepts can be marked as top concepts. This helps to provide ``efficient access''~\cite{Isaac2009} and simplifies orientation in the vocabulary. 
% Function definition
We provide a function \textit{otc} that iterates over all \texttt{skos:ConceptScheme}s in $V$ and collects those that do not occur in relations established by the properties \texttt{skos:hasTopConcept} or \texttt{skos:topConceptOf}.

\paragraph{QI 11: Top Concept Having Broader Concepts}
% Origin and rationale
Allemang et al. \cite{Allemang2011} propose to ``not indicate any concepts internal to the tree as top concepts'', i. e. top concepts should not have broader concepts. 
% Function definition
The function \textit{tchbc} iterates over all \texttt{skos:TopConcept}s defined in $V$ and collects those that are related to a resource by the \texttt{skos:broader} property.

@@

This section identifies issues related to a vocabulary's interaction with other datasets in the LOD cloud.

\paragraph{QI 12: Missing In-Links}
% Origin and rationale
A periodically updated version of a diagram visualizing ``the data sets in the LOD cloud as well as their interlinkage relationships'' is publised online\footnote{\url{http://www4.wiwiss.fu-berlin.de/lodcloud/state/}}. In these diagrams, DBpedia is shown as the most intensely linked resource, i.e. a large number of vocabularies link to resources in that dataset. This indicates that a dataset with a large number of other datasets referencing it, can be assumed to be of great value for the community, adding cohesion to the LOD cloud. Accordingly, estimating the number of In-Links of SKOS concepts defined in a vocabulary (henceforth called the in-degree of a concept) gives some hint about how established or accepted a controlled vocabulary is. A method to carry out these estimations would be utilizing existing Linked Data indices like, e.g., Sindice\footnote{\url{http://sindice.com/}} that provide a SPARQL endpoint. 
% Function definition
The function \textit{mil} estimates the number of In-Links for a conceptual resource, by iterating over all elements in $C$ and querying the Sindice SPARQL endpoint for triples containing the concept's URI in the object part. Those concepts having an empty query result, are missing any In-Links and are thus included in the result set of this issue.

\paragraph{QI 13: Missing Out-Links}
% Origin and rationale
The Web of Data consists of globally interconnected datasets, ``enabling seamless connections between data sets''\cite{Heath2011}. These links can be established between various SKOS vocabularies by Out-Links, i.e., references from conceptual resources in the ``source vocabulary'' $V$ to resources identified by an HTTP URI and available at a different host. Similar to QI 12, this issue identifies the set of all conceptual resources that lack such Out-Links. 
% Function definition
We provide the function \textit{mol} that iterates over all elements $c \in C$ and investigates the host part of all related resources. If none of these host names differ from the host part of $c$, it has no Out-Links and is thus included into the resources identified by this issue.

\subsubsection{QI 14: Link Target Unavailability} \todo{BH}{should we just call this "Broken Links" because that's essentially what it's all about}
% Origin and rationale
This criterion identifies a set of all resources that have been identified as unavailable, i.e. resolving their URI leads to an erroneous HTTP response or no response at all. An erroneous HTTP response in that case can be defined as a response code (after a possible redirection) other than 200. Just as in the ``traditional'' Web, these ``broken links'' hinder information gathering and should be avoided. 
% Function definition
Calculation of this criterion is performed by iterating over all resources in $IR$, performing an HTTP request and include those resources in the result set that are unavailable.

% OTHER CRITERIA

\subsection{Other Criteria}

\subsubsection{Undefined SKOS Resource Usage}
Some vocabularies reference resources in the SKOS namespace that cannot be resolved. A common reason for this is, that there seems to exist a misunderstanding about the correct way of introducing new classes or properties in an OWL document. In some cases, vocabularies ``invent'' new terms in the SKOS namespace in order to meet certain requirements of the publishing organizations that cannot (yet) be expressed within the SKOS schema. Another reason for the presence of unresolvable resource references in the SKOS namespace is that the vocabulary might be outdated. With the SKOS W3C recommendation maturing, some properties have been removed from the current version of the schema. Vocabularies using these ``deprecated'' properties hence contain parts of their information in a non ``standard"-compliant way, having negative effects when, e.g., crawling vocabularies to gather specific information. A method to identify these undefined SKOS resources is to iterate over all resources in $IR$, including (i) those resources in the resulting set that are contained in the list of deprecated resources or (ii) have an URI in the SKOS namespace but are not defined in the SKOS ontology.

>>>>

@@

@@

@@

% Origin and rationale
The ISO/DIS 25964-1 standard \cite{ISO25964-1:2011} suggests that terms that share a common broader term but do not have an overlapping meaning, should not be related associatively. This is advocated by Hedden~\cite{Hedden2010} and Aitchison et al.~\cite{,Aitchison2000} who point out ``the risk that thesaurus compilers may overload the thesaurus with valueless relationships'', having a negative effect on precision.
% Function defintion
The function \textit{var} identifies a set of pairs of conceptual resources that share the same broader or narrower concept while also being associatively related by the property \texttt{skos:related}.

\paragraph{QI 9: Solely Transitively Related Concepts}

% Origin and rationale
Two concepts that are explicitly related by \texttt{skos:broaderTransitive} and/or \texttt{skos:narrowerTransitive} can be regarded a quality issue because, according to \cite{SkosReference2008}, they are ``not used to make assertions''. Transitive hierarchical relations in SKOS are meant to be inferred by the vocabulary consumer, which is reflected in the SKOS schema by, e.g., \texttt{skos:broader} being a subproperty of \texttt{skos:broaderTransitive}. 
% Function definition
We encode this issue in a function \textit{strc} that iterates over all concept pairs $C \times C$ that are directly related by the mentioned transitive properties, not by one of their subproperties.

% TODO CM: it is unclear what "mentioned" means in this context...

\paragraph{QI 10: Omitted Top Concepts}

% Origin and rationale
The SKOS language provides concept schemes, which are a facility for grouping related concepts. This helps to provide ``efficient access''~\cite{Isaac2009} and simplifies orientation in the vocabulary. In order to provide entry points to such a group of concepts, one or more concepts can be marked as top concepts.  
% Function definition
We provide a function \textit{otc} that iterates over all \texttt{skos:ConceptScheme}s in $V$ and collects those that do not occur in relations established by the properties \texttt{skos:hasTopConcept} or \texttt{skos:topConceptOf}.

\paragraph{QI 11: Top Concept Having Broader Concepts}

% Origin and rationale
Allemang et al. \cite{Allemang2011} propose to ``not indicate any concepts internal to the tree as top concepts'', which means that top concepts should not have broader concepts. 
% Function definition
The function \textit{tchbc} iterates over all \texttt{skos:TopConcept}s defined in $V$ and collects those that are related to a resource by the \texttt{skos:broader} property.

@@

\paragraph{QI 12: Missing In-Links}

% Origin and rationale
SKOS concepts become linkable resources, when vocabularies are being published on the Web. Estimating the number of in-links and identifying the concepts without any in-links, can be an important indicator for the importance of a concept. The required in-links for such an estimation can be obtained from existing Linked Data indices like, e.g., Sindice\footnote{\url{http://sindice.com/}}. 
% Function definition
The function \textit{mil} estimates the number of in-links referencing a concept, by iterating over all elements in $C$ and querying the Sindice SPARQL endpoint for triples containing the concept's URI in the object part. Empty query results are indicator for missing  in-links and are thus included in the result set of this function.

\paragraph{QI 13: Missing Out-Links}

% Origin and rationale
In the Linked Data context SKOS concepts should be linked with other related concepts on the Web, ``enabling seamless connections between data sets''\cite{Heath2011}. Similar to \emph{Missing In-Links}, this issue identifies the set of all conceptual resources that have no out-links. 
% Function definition
We provide the function \textit{mol} that iterates over all elements $c \in C$ and investigates \todo{CM}{mapping relations?} the host part of all related resource URIs. If none of these host names differ from the host part of $c$'s URI, it has no out-links and is thus included in the resources identified by this issue.

\paragraph{QI 14: Broken Links}

% Origin and rationale
This issue identifies vocabulary resources that return HTTP error responses or no response when being dereferenced. An erroneous HTTP response in that case can be defined as a response code other than 200 when taking possible redirections into account. Just as in the ``document'' Web, these ``broken links'' hinder navigability also in the Linked Data Web and and should therefore be avoided. 
% Function definition
Broken links are identified by a function $bl$, which iterates over all resources in $IR$, dereferences each resource, follows possible redirects, and includes unavailable resources in the result set.

\paragraph{QI 15: Undefined SKOS Resources}

% Origin and rationale
SKOS defines a set of classes and properties, hence resources within the namespace \url{http://www.w3.org/2004/02/skos/core#}, for building controlled vocabularies. However, some vocabularies use resources from within this namespace, which are unresolvable for two main reasons: vocabulary creators ``invented'' new terms within the SKOS namespace instead of introducing them in a separate namespace, or they use ``deprecated'' SKOS properties.
% Function definition
In order to identify such undefined SKOS resources, we define a function $usr$  that iterates over all resources in $IR$ and returns those resources (i) that are contained in the list of deprecated resources or (ii) are identified by a URI in the SKOS namespace but are not defined in the current version of the SKOS ontology.
